---
title: "Download and visualize ERS data"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: false
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{RstoxStrata}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE, comment = "#>")
```

<!-- # Summary -->

<!-- Blap -->

Packages required to run the examples:

```{r}
library(sf)
library(dplyr)
library(tidyr)
library(tibble)
library(RstoxUtils)
library(ggOceanMaps)
library(cowplot)
```

# Introduction

The Directorate of Fisheries in Norway (FDir) collects and distributes data on [exact positions and times of commercial fishing events by the Norwegian fleet](https://www.fiskeridir.no/english/Fisheries/reporting-systems-and-innovation/electronic-reporting-ers-and-position-reporting-vms-in-norwegian-fisheries) for vessels >15 m. These data, called ERS for Electronic Registration System, are available on the [FDir website](https://www.fiskeridir.no/statistikk-tall-og-analyse/data-og-statistikk-om-yrkesfiske/apne-data-elektronisk-rapportering-ers). These data are currently only available in Norwegian compressed in large unsearchable zip files. The `downloadERS()` and `extractERS()` functions aid in efficient handling of these data. 

## Download ERS data

## Process ERS data



